---
title: "Nested Positions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nested-positions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(ggdibbler)
library(patchwork)
```

### Position Nesting
One of the most powerful features of `ggdibbler` is the ability to do nested positions. That is, you can control the overplotting in the original plot, *and* the over plotting caused by resampling using separate positions. This can create some interested graphics. 

Below is a grid of box plots. The first column is the original ggplot we are going to add uncertainty to, the second and third column is the effect of the overplotting being managed with an nested identity (with decreased alpha) and dodge positions respectively. The syntax of the nested positions is `mainposition_nestedposition`, where the main position is what was in the original ggplot, and the nested position manages the overplotting caused by sampling from the distribution. 

```{r, fig.width=16, fig.height=16}
# ggplot IDENTITY
p1 <- ggplot(mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), 
                  position = "identity", alpha=0.7)+
  theme(legend.position="none", aspect.ratio = 1) +
  ggtitle("ggplot2: identity")
# ggdibbler identity
p2 <- ggplot(uncertain_mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), alpha=0.1,
                  position = "identity_identity")+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("identity_identity")

p3 <- ggplot(uncertain_mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), alpha=0.7,
                  position = "identity_dodge")+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("identity_dodge")


# ggplot dodge
p5 <- ggplot(mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), 
                  position = position_dodge(preserve="single"))+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("ggplot: dodge")

p6 <- ggplot(uncertain_mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), alpha=0.1,
                  position = "dodge_identity")+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("dodge_identity")

p7 <- ggplot(uncertain_mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), 
                  position = "dodge_dodge")+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("dodge_dodge")


# ggplot stack
p9 <- ggplot(mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), 
                  position = "stack")+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("ggplot2: stack")


p10 <- ggplot(uncertain_mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv), alpha=0.1,
                  position = "stack_identity")+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("stack_identity")

p11 <- ggplot(uncertain_mpg, aes(class)) + 
  geom_bar_sample(aes(fill = drv),
                  position = "stack_dodge")+
  theme(legend.position="none", aspect.ratio = 1)+
  ggtitle("stack_dodge")


(p1 | p2 | p3 ) / (p5 | p6 | p7) / (p9 | p10 | p11)
```
